\bf Effects without a Cause: The Search for Demographic Anomalies

### 2020-01-17 MANUSCRIPT/manuscript.Rmd
Display items finished.
To do:
- Perhaps an example in the Methods section?
- A little more text in the Results section
- A conclusion
- Rewrite the Abstract

library(knitr)
read_chunk('../R/SCRIPTS/000-Libraries.R')
read_chunk('../R/SCRIPTS/001-DataLoad.R')
read_chunk('../R/SCRIPTS/002-StatesAnalysis.R')
```

@@

<!-- # ```{r StatesAnalysis, include=FALSE} -->
<!-- # ``` -->

@@

where $\pi(B)=\sum_{i=o}^{inf} \pi_iB^i$. The identification of outliers then involves a three step process to (1) identify all potential outliers, $t_j$ and $L_j (B)$, (2) joint estimates of model parameters and outlier effects are computed to identify potentially spurious outliers, and (3) the outliers and effects are re-estimated without spurious outliers. Of importance here is step 2, which relies on some critical value above which an outlier at time point m is considered spurious. Based on Chen and Liu’s [-@chen1993joint] recommendation, we set a critical value of 3.5 which generally minimizes the possibility of Type I errors or false positive outliers. Outliers are reported using a simple t-statistic.

@@

**\autoref{sumtable}** reports the overall stuff.

@@

 anom_mort <- read_rds("../R/DATA-PROCESSED/mortality_anomalies") %>%
   na.omit %>%
   mutate(component = "Mortality")

@@

 anom <- rbind(anom_mort, anom_fert)

 types <- anom %>%
   group_by(type, component) %>%
   dplyr::summarise(n = n(),
                    states =  n_distinct(location)) %>%
   ungroup() %>%
   gather(test, stat_, n:states) %>%
   mutate(variable = paste0(component, "_", test)) %>%
   ungroup() %>%
   dplyr::select(-component, -test) %>%
   spread(variable, stat_)

 kable(types,
       "latex",
       booktabs = T,
       col.names = c( "Type","n","States","n","States"),
       caption = "\\textbf{Summary by Anomaly Type and Demographic Component.} n refers to the number of anomolies for each type and demogrpahic component, States refers to the number of unique states exhibiting that specific Type-demogrpahic component. \\label{sumtable}") %>%
   kable_styling(position = "center") %>%
   add_header_above(c(" "=1, "Fertility" = 2, "Mortality" = 2))

>>>>

library(knitr)
read_chunk('../R/SCRIPTS/000-Libraries.R')
read_chunk('../R/SCRIPTS/001-DataLoad.R')
read_chunk('../R/SCRIPTS/002-StatesAnalysis.R')
read_chunk('../R/SCRIPTS/003-SummaryStats.R')
```

@@

```{r SumStats, include=FALSE}
```

@@

where $\pi(B)=\sum_{i=o}^{inf} \pi_iB^i$. The identification of outliers then involves a three step process to (1) identify all potential outliers, $t_j$ and $L_j (B)$, (2) joint estimates of model parameters and outlier effects are computed to identify potentially spurious outliers, and (3) the outliers and effects are re-estimated without spurious outliers. Of importance here is step 2, which relies on some critical value above which an outlier at time point m is considered spurious. Based on Chen and Liu’s [-@chen1993joint] recommendation, we set a critical value of 3.5 which generally minimizes the possibility of Type I errors or false positive outliers.

We report outliers using a simple t-statistic and report the size of the outlier, or its impact on the time series, by subtracting $\hat{y}$, the adjusted time series without outliers removed, from $y$, the original, unadjusted time series.

@@

@@

sums2 <- sums %>%
  dplyr::select(percent_anom = `\\% of Total`, everything())
kable(sums,
       "latex",
       booktabs = T,
       format.args = list(big.mark = ","),
       escape = FALSE,
       caption = "\\textbf{Summary by Demographic Component.}  refers to the number of anomolies for each type and demographic component, States refers to the number of unique states exhibiting that specific Type-demographic component. \\label{sumtable}") %>%
   kable_styling(position = "center")

**\autoref{sumtable}** report the overall number of anomalies we detect of each type for births and deaths and some summary statistics across all anomaly types. We find considerably more mortality anomalies (n = `r paste(sums[2,2])`) than fertility anomalies (n = `r paste(sums[1,2])`). This is likely because mortality is more easily subject to instantaneous change compared with fertility. Changes in fertility should theoretically manifest approximately 9-months after change.

We also find more anomalous deaths (`r paste(prettyNum(sums[2,6], digits = 0, big.mark = ","))`) than anomalous births (`r paste(prettyNum(sums[1,6], digits = 0, big.mark = ","))`). Fertility anomalies overwhelmingly tended toward lower fertility, with only a few fertility anomalies yielding more births. Conversely, mortality anomalies tended to be more evenly split between positive and negative anomalies, but tended toward more deaths rather than less deaths. These anomalous deaths and births account for `r sums2$percent_anom[2]`% and `r sums2$percent_anom[1]`% over the entire time series, respectively.

**\autoref{fig:summap}** maps these results. Three states exhibited neither mortality nor fertility anomalies: Alaska, North Dakota, and South Dakota. Another nine states exhibited just a single anomaly (DE, DC, ID, IL, KS, NV, NM, UT, and WY). Both New York and Massachusetts exhibited the most anomalies with nine.

```{r AnomalyMap, echo= FALSE, message = FALSE, warning = FALSE, fig.height=7, fig.cap= paste("\\textbf{Summary of Anomalies by type, demographic component, and State.} We observe much fewer fertility anomalies than mortality anomalies. \\label{fig:summap}")}

data(state_laea)
anom_mort <- read_rds("../R/DATA-PROCESSED/mortality_anomalies") %>%
  na.omit %>%
  mutate(component = "Mortality")

@@

anom_mort2 <- rbind(anom_mort, anom_fert) %>%
  group_by(type, location, component) %>%
  dplyr::summarise(Anomalies =  n()) %>%
  mutate(ident = paste0(component, " - ", type),
         GEOID = substr(location, 2,3)) %>%
  ungroup() %>%
  dplyr::select(GEOID, Anomalies, ident) %>%
  spread(ident, Anomalies) %>%
    mutate_all(funs(replace_na(.,0)))

states <- left_join(state_laea, anom_mort2) %>%
  gather(ident, Anomalies, `Fertility - AO`:`Mortality - TC`) %>%
  mutate_all(funs(replace_na(.,0))) %>%
  mutate(Anomalies = factor(Anomalies))

states$ident2 <- factor(
  states$ident,
  levels = c("Mortality - AO", "Fertility - AO", 
             "Mortality - TC", "Fertility - TC", 
             "Mortality - LS", "Fertility - LS")
)

colorpal <- c("#ffffff", "#ffffcc", "#a1dab4", "#41b6c4", "#2c7fb8", "#253494")

  ggplot(data=states) +
  geom_sf(aes(fill = Anomalies),  lwd = 1/2) +
  theme_bw() +
  theme(axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(),
        legend.position="bottom"
  )+
  guides(fill=guide_legend(nrow=1,byrow=TRUE)) +
  facet_wrap(~ident2, ncol=2) +
    scale_fill_manual(name = "# of Anomalies", values = colorpal) +

  NULL
```

### 2020-01-17 MANUSCRIPT/manuscript.tex
Display items finished.
To do:
- Perhaps an example in the Methods section?
- A little more text in the Results section
- A conclusion
- Rewrite the Abstract

where \(\pi(B)=\sum_{i=o}^{inf} \pi_iB^i\). The identification of
outliers then involves a three step process to (1) identify all
potential outliers, \(t_j\) and \(L_j (B)\), (2) joint estimates of
model parameters and outlier effects are computed to identify
potentially spurious outliers, and (3) the outliers and effects are
re-estimated without spurious outliers. Of importance here is step 2,
which relies on some critical value above which an outlier at time point
m is considered spurious. Based on Chen and Liu's
\citeyearpar{chen1993joint} recommendation, we set a critical value of
3.5 which generally minimizes the possibility of Type I errors or false
positive outliers. Outliers are reported using a simple t-statistic.

@@

>>>>

where \(\pi(B)=\sum_{i=o}^{inf} \pi_iB^i\). The identification of
outliers then involves a three step process to (1) identify all
potential outliers, \(t_j\) and \(L_j (B)\), (2) joint estimates of
model parameters and outlier effects are computed to identify
potentially spurious outliers, and (3) the outliers and effects are
re-estimated without spurious outliers. Of importance here is step 2,
which relies on some critical value above which an outlier at time point
m is considered spurious. Based on Chen and Liu's
\citeyearpar{chen1993joint} recommendation, we set a critical value of
3.5 which generally minimizes the possibility of Type I errors or false
positive outliers.

We report outliers using a simple t-statistic and report the size of the
outlier, or its impact on the time series, by subtracting \(\hat{y}\),
the adjusted time series without outliers removed, from \(y\), the
original, unadjusted time series.

@@

\begin{table}

\caption{\label{tab:unnamed-chunk-1}\textbf{Summary by Demographic Component.}  refers to the number of anomolies for each type and demographic component, States refers to the number of unique states exhibiting that specific Type-demographic component. \label{sumtable}}
\centering
\begin{tabular}[t]{lrllrrr}
\toprule
Component & Anomalies & $y$ & $\hat{y}$ & $y-\hat{y}$ & $|y-\hat{y}|$ & \% of Total\\
\midrule
Fertility & 22 & 25.48M & 25.68M & -202,217 & 226,476 & 0.889\\
Mortality & 156 & 43.57M & 43.52M & 48,535 & 627,364 & 1.440\\
\bottomrule
\end{tabular}
\end{table}

\textbf{\autoref{sumtable}} report the overall number of anomalies we
detect of each type for births and deaths and some summary statistics
across all anomaly types. We find considerably more mortality anomalies
(n = 156) than fertility anomalies (n = 22). This is likely because
mortality is more easily subject to instantaneous change compared with
fertility. Changes in fertility should theoretically manifest
approximately 9-months after change.

We also find more anomalous deaths (627,364) than anomalous births
(226,476). Fertility anomalies overwhelmingly tended toward lower
fertility, with only a few fertility anomalies yielding more births.
Conversely, mortality anomalies tended to be more evenly split between
positive and negative anomalies, but tended toward more deaths rather
than less deaths. These anomalous deaths and births account for 1.44\%
and 0.889\% over the entire time series, respectively.

\textbf{\autoref{fig:summap}} maps these results. Three states exhibited
neither mortality nor fertility anomalies: Alaska, North Dakota, and
South Dakota. Another nine states exhibited just a single anomaly (DE,
DC, ID, IL, KS, NV, NM, UT, and WY). Both New York and Massachusetts
exhibited the most anomalies with nine.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/AnomalyMap-1.pdf}
\caption{\textbf{Summary of Anomalies by type, demographic component, and State.}
We observe much fewer fertility anomalies than mortality anomalies.
\label{fig:summap}}
\end{figure}
